---
title: "DADA2 + Phyloseq Pipeline for 16S rDNA Landfill Microcosms"
author: "Gavin McNicol / ecoφlab"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: readable
params:
  project: "landfill_microcosms"   # change per run
  silva_path: "data/SILVA_SSU_r138_2019.RData"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE)
```

# Setup + Package Loading

```{r dada2, eval = F}
# library(devtools)
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("dada2")
# BiocManager::install("phyloseq")
# BiocManager::install("DECIPHER")
# BiocManager::install("phangorn")
# BiocManager::install("DESeq2")
# BiocManager::install("microbiome")
```

```{r other-package}
library(dada2)
library(phyloseq)
library(DECIPHER)
library(phangorn)
library(DESeq2)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(vegan)
library(microbiome)
library(doParallel)
library(viridis)

registerDoParallel(cores = parallel::detectCores() - 1)  # parallel processing
```

# Parameterized File Setup

```{r file-paths}
# Define paths (easy to change)
data_dir <- "data"
fwd_dir <- file.path(data_dir, "forward")
rev_dir <- file.path(data_dir, "reverse")
filt_dir <- file.path(data_dir, "filtered")

dir.create(filt_dir, showWarnings = FALSE, recursive = TRUE)

# Get sorted fastq files
fnFs <- sort(list.files(fwd_dir, pattern = "_R1", full.names = TRUE))
fnRs <- sort(list.files(rev_dir, pattern = "_R2", full.names = TRUE))

sample_names <- sapply(strsplit(basename(fnFs), "_"), `[`, 1)
names(fnFs) <- sample_names
names(fnRs) <- sample_names

cat("Found", length(fnFs), "samples\n")
```

# Create/Import metadata file

```{r load-metadata}
# Option 1: Read from CSV (recommended)
# all_metadata <- read.csv("data/metadata.csv", row.names = 1, stringsAsFactors = FALSE)

# Option 2: If you need to create one from scratch (example)
all_metadata <- data.frame(
  sample = sample_names,
  site = c("WW", "BP", "SW", "EW", "NW", "SP"),
  # treatment = c("Control", "Fe(OH)3", ...),
  status = "Live",
  replicate = 1,
  row.names = sample_names
)

# Important: Make sure row names match exactly
rownames(all_metadata) <- all_metadata$sample   # or whatever your ID column is

# Check alignment
cat("Do row names match the ASV table?\n")
cat(all(rownames(seqtab.nochim) == rownames(all_metadata)), "\n")
```


# Quality Check + Filtering (Modern `filterAndTrim`)

```{r quality-filter}
# Quality profiles 
plotQualityProfile(fnFs[1:min(6, length(fnFs))])
plotQualityProfile(fnRs[1:min(6, length(fnRs))])

# Filter and trim
filtFs <- file.path(filt_dir, basename(fnFs))
filtRs <- file.path(filt_dir, basename(fnRs))

out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs,
                     trimLeft = c(19, 20),      # remove primers
                     truncLen = c(150, 130),    # ← updated to 130 to be stricter on R2
                     maxN = 0,
                     maxEE = c(2, 2),           # or try c(2, 3) if too many reads lost
                     truncQ = 2,
                     compress = TRUE,
                     multithread = TRUE)

head(out)

# Quality profiles again
plotQualityProfile(filtFs[1:min(6, length(filtFs))])
plotQualityProfile(filtRs[1:min(6, length(filtRs))])
```


# Error Learning + DADA2 (Better Practice)

```{r learn-errors, cache=TRUE}
# Learn error rates (this can take a while — verbose provides updates)
cat("=== Learning error rates for Forward reads ===\n")
errF <- learnErrors(filtFs, 
                    multithread = TRUE, 
                    randomize = TRUE, 
                    nbases = 1e8,      # adjust based on your data size
                    verbose = TRUE)    # ← This gives iterative progress updates

cat("\n=== Learning error rates for Reverse reads ===\n")
errR <- learnErrors(filtRs, 
                    multithread = TRUE, 
                    randomize = TRUE, 
                    nbases = 1e8,
                    verbose = TRUE)

# Plot the learned error models
plotErrors(errF, nominalQ = TRUE) + ggtitle("Forward Error Rates")
plotErrors(errR, nominalQ = TRUE) + ggtitle("Reverse Error Rates")
```


```{r de-replicate}
# Dereplicate
derepFs <- derepFastq(filtFs, verbose = TRUE)
derepRs <- derepFastq(filtRs, verbose = TRUE)

# Assign sample names (important!)
names(derepFs) <- sample_names
names(derepRs) <- sample_names

cat("=== Running DADA on Forward reads ===\n")
dadaFs <- dada(derepFs, err = errF, pool = TRUE, multithread = TRUE, verbose = TRUE)

cat("\n=== Running DADA on Reverse reads ===\n")
dadaRs <- dada(derepRs, err = errR, pool = TRUE, multithread = TRUE, verbose = TRUE)
```


# Merge Pairs + Chimera Remova

 THIS DID NOT WORK -- Something to do with earlier PEAR merge?
```{r merge-chimera, cache=TRUE, eval = F}
# # ====================== MERGE PAIRS ======================
# cat("=== Merging paired reads ===\n")
# mergers <- mergePairs(dadaFs, derepFs, dadaRs, derepRs, 
#                       verbose = TRUE,
#                       minOverlap = 8,          # was 6 in old script
#                       maxMismatch = 0,
#                       trimOverhang = TRUE)
# 
# # Optional: also try a more lenient version if needed
# # mergers2 <- mergePairs(dadaFs, derepFs, dadaRs, derepRs, 
# #                        verbose = TRUE, minOverlap = 6)
# 
# # ====================== SEQUENCE TABLE ======================
# cat("=== Making sequence table ===\n")
# seqtab <- makeSequenceTable(mergers)
# 
# # Remove chimeras
# cat("=== Removing chimeras ===\n")
# seqtab.nochim <- removeBimeraDenovo(seqtab, 
#                                     method = "consensus", 
#                                     multithread = TRUE, 
#                                     verbose = TRUE)
# 
# # Summary of chimera removal
# cat("\nSummary:\n")
# cat("Input sequences:", sum(seqtab), "\n")
# cat("After chimera removal:", sum(seqtab.nochim), "\n")
# cat("Percent retained:", round(100 * sum(seqtab.nochim) / sum(seqtab), 1), "%\n")
# 
# # Track reads through pipeline
# getN <- function(x) sum(getUniques(x))
# track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), 
#                sapply(mergers, getN), rowSums(seqtab.nochim))
# colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nochim")
# rownames(track) <- sample_names
# head(track)
```

Simpler merge -- did not work either 
```{r forward-only-reads}
# # Use only Forward reads (much higher yield)
# seqtab <- makeSequenceTable(dadaFs)                    # ← only Forward
# seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", 
#                                     multithread=TRUE, verbose=TRUE)
# 
# # Summary of chimera removal
# cat("\nSummary:\n")
# cat("Input sequences:", sum(seqtab), "\n")
# cat("After chimera removal:", sum(seqtab.nochim), "\n")
# cat("Percent retained:", round(100 * sum(seqtab.nochim) / sum(seqtab), 1), "%\n")
# 
# # Track reads through pipeline
# getN <- function(x) sum(getUniques(x))
# track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN),
#                sapply(mergers, getN), rowSums(seqtab.nochim))
# colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nochim")
# rownames(track) <- sample_names
# head(track)
```

# Forward only seqtab

```{r seqtab-chimera-forward-only, cache=TRUE}
# Use only Forward reads (best option given poor merging)
cat("=== Making sequence table from Forward reads only ===\n")
seqtab <- makeSequenceTable(dadaFs)

cat("=== Removing chimeras ===\n")
seqtab.nochim <- removeBimeraDenovo(seqtab, 
                                    method = "consensus", 
                                    multithread = TRUE, 
                                    verbose = TRUE)

# Summary
cat("\nSummary:\n")
cat("Input sequences:", sum(seqtab), "\n")
cat("After chimera removal:", sum(seqtab.nochim), "\n")
cat("Percent retained:", round(100 * sum(seqtab.nochim) / sum(seqtab), 1), "%\n")

```

# Create clean read tracking table for Table S3 (Supplemental)

```{r}
# === Create clean tracking table for Table S3 ===
getN <- function(x) sum(getUniques(x))

track <- data.frame(
  sample     = sample_names,
  input      = out[, "reads.in"],
  filtered   = out[, "reads.out"],
  denoisedF  = sapply(dadaFs, getN),
  seqtab     = rowSums(seqtab),
  nochim     = rowSums(seqtab.nochim)
)

# Add % retained at each step
track$filtered_pct <- round(track$filtered / track$input * 100, 1)
track$seqtab_pct   <- round(track$seqtab   / track$filtered * 100, 1)
track$nochim_pct   <- round(track$nochim   / track$seqtab * 100, 1)
track$final_pct    <- round(track$nochim   / track$input * 100, 1)

# Nice column order
track <- track[, c("sample", "input", "filtered", "filtered_pct", 
                   "denoisedF", "seqtab", "seqtab_pct", 
                   "nochim", "nochim_pct", "final_pct")]

print(track)

# Save for supplemental
write_csv(track, "output/Table_S3_DADA2_read_tracking.csv")
```




# Taxonomy (DECIPHER)

get SILVA dataset from: 
url = "https://drive.google.com/uc?export=download&id=1w3wdSCpSihntWkbP_zvXz7r3s-tNB8DV",

```{r}
load("data/SILVA_SSU_r138.2_v2.RData")
# The object is usually named `trainingSet`
ls()  # should show trainingSet
```


```{r taxonomy-decipher, cache=TRUE}
library(DECIPHER)

# Prepare sequences
dna <- DNAStringSet(getSequences(seqtab.nochim))

cat("=== Running DECIPHER IdTaxa ===\n")
ids <- IdTaxa(dna, 
              trainingSet, 
              strand = "top", 
              threshold = 60, 
              processors = NULL, 
              verbose = TRUE)

# Convert to tax table matrix
ranks <- c("domain", "phylum", "class", "order", "family", "genus", "species")
taxid <- t(sapply(ids, function(x) {
  m <- match(ranks, x$rank)
  taxa <- x$taxon[m]
  taxa[startsWith(taxa, "unclassified_")] <- NA
  taxa
}))

colnames(taxid) <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus", "Species")
rownames(taxid) <- getSequences(seqtab.nochim)

# Save outputs
saveRDS(seqtab.nochim, "output/seqtab_nochim_fwd.rds")
saveRDS(taxid, "output/taxid_fwd.rds")

cat("Taxonomy assignment complete. Assigned to", ncol(taxid), "ranks.\n")
head(taxid[, 1:7])  # preview
```

# Make phyloseq object

```{r make-phyloseq, cache=TRUE}
# Create phyloseq object (forward-only)
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = FALSE),
               tax_table(taxid),
               sample_data(all_metadata))   # ← load your metadata here first

# Basic summary
cat("Phyloseq object created with", nsamples(ps), "samples and", ntaxa(ps), "ASVs\n")
cat("Taxonomic ranks:", colnames(tax_table(ps)), "\n")

# Save it
saveRDS(ps, "output/nachusa-2022-fwd-nochim.rds")

# Optional: Quick summary at family/genus level
ps_family <- tax_glom(ps, taxrank = "Family")
ps_genus  <- tax_glom(ps, taxrank = "Genus")

cat("Family-level ASVs:", ntaxa(ps_family), "\n")
cat("Genus-level ASVs:", ntaxa(ps_genus), "\n")
```
# Final Phyloseq Object

```{r finalize-and-summarize, cache=TRUE}
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows = FALSE),
               tax_table(taxid),
               sample_data(all_metadata))

# Save everything
# saveRDS(ps, "output/ps-nachusa-2022-fwd-nochim-final.rds")
# saveRDS(ps_family, "output/ps-family.rds")
# saveRDS(ps_genus, "output/ps-genus.rds")

ps <- readRDS("output/ps-nachusa-2022-fwd-nochim-final.rds")
ps_family <- readRDS("output/ps-family.rds")
ps_genus <- readRDS("output/ps-genus.rds")


# === FAMILY LEVEL - Top abundances PER SAMPLE ===
cat("\n=== Top 10 Most Abundant Families per Sample ===\n")

family_abund_by_sample <- tax_glom(ps, taxrank = "Family") %>%
  transform_sample_counts(function(x) 100 * x / sum(x)) %>%   # relative abundance per sample
  psmelt() %>%
  group_by(Sample, Family) %>%
  summarise(Rel_Abund = sum(Abundance), .groups = "drop") %>%
  group_by(Sample) %>%
  arrange(Sample, desc(Rel_Abund)) %>%
  slice_head(n = 10) %>%                    # Change to 15 or 20 if you want more
  ungroup()

print(family_abund_by_sample)

# Optional: Save as CSV
write_csv(family_abund_by_sample, "output/top_families_per_sample.csv")

# === GENUS LEVEL - Top abundances PER SAMPLE ===
cat("\n=== Top 10 Most Abundant Genera per Sample ===\n")

genus_abund_by_sample <- tax_glom(ps, taxrank = "Genus") %>%
  transform_sample_counts(function(x) 100 * x / sum(x)) %>%
  psmelt() %>%
  group_by(Sample, Genus) %>%
  summarise(Rel_Abund = sum(Abundance), .groups = "drop") %>%
  group_by(Sample) %>%
  arrange(Sample, desc(Rel_Abund)) %>%
  slice_head(n = 10) %>%
  ungroup()

print(genus_abund_by_sample)

# Optional: Save as CSV
write_csv(genus_abund_by_sample, "output/top_genera_per_sample.csv")
```

# Alpha Diversity Table + Plot

```{r alpha-diversity, fig.width=10, fig.height=6}
# Alpha diversity metrics
alpha_div <- estimate_richness(ps, measures = c("Observed", "Shannon", "Simpson")) %>%
  rownames_to_column("SampleID") %>%
  left_join(all_metadata %>% rownames_to_column("SampleID"), by = "SampleID")

cat("=== Alpha Diversity Summary ===\n")
print(alpha_div)

# Save table
write.csv(alpha_div, "output/alpha_diversity.csv", row.names = FALSE)

# Plot
p_alpha <- alpha_div %>%
  pivot_longer(cols = c(Observed, Shannon, Simpson),
               names_to = "Metric", values_to = "Value") %>%
  ggplot(aes(x = SampleID, y = Value, fill = SampleID)) +   # change fill to your main factor
  geom_col() +
  facet_wrap(~ Metric, scales = "free_y") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Alpha Diversity - Landfill Microcosms",
       x = "", y = "Diversity Metric")

print(p_alpha)
ggsave("figures/alpha_diversity_plot.png", p_alpha, width = 10, height = 6, dpi = 300)
```


# Relative Abundance Barplots (Family & Genus)

```{r rel-abund-barplots-improved, fig.width=12, fig.height=7}
add_other_family <- function(ps_rel, taxrank = "Family", threshold = 2) {
  psmelt(ps_rel) %>%
    group_by(Sample) %>%
    mutate(Taxon = ifelse(Abundance < threshold, "Other", 
                          .data[[taxrank]])) %>%     # safe dynamic column
    group_by(Sample, Taxon) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop")
}

# Family level
ps_fam_rel <- tax_glom(ps, "Family") %>% 
  transform_sample_counts(function(x) 100 * x / sum(x))

p_family <- add_other_family(ps_fam_rel, threshold = 2) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Taxon)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_viridis_d(option = "turbo") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        legend.text = element_text(size = 7.5)) +
  labs(title = "Family-Level Relative Abundance",
       y = "Relative Abundance (%)", x = "")

print(p_family)
ggsave("figures/family_rel_abund_100pct.png", p_family, width = 12, height = 7, dpi = 300)

# Genus level (more aggressive "Other" because there are more taxa)
add_other_genus <- function(ps_rel, taxrank = "Genus", threshold = 2) {
  psmelt(ps_rel) %>%
    group_by(Sample) %>%
    mutate(Taxon = ifelse(Abundance < threshold, "Other", 
                          .data[[taxrank]])) %>%     # safe dynamic column
    group_by(Sample, Taxon) %>%
    summarise(Abundance = sum(Abundance), .groups = "drop")
}

ps_gen_rel <- tax_glom(ps, "Genus") %>% 
  transform_sample_counts(function(x) 100 * x / sum(x))

p_genus <- add_other_genus(ps_gen_rel, taxrank = "Genus", threshold = 3) %>%
  ggplot(aes(x = Sample, y = Abundance, fill = Taxon)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_viridis_d(option = "plasma") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "right",
        legend.text = element_text(size = 7)) +
  labs(title = "Genus-Level Relative Abundance",
       y = "Relative Abundance (%)", x = "")

print(p_genus)
ggsave("figures/genus_rel_abund_100pct.png", p_genus, width = 12, height = 7, dpi = 300)
```

# Beta-diversity (between plots)

```{r beta-diversity-ordination, fig.width=10, fig.height=5}
library(vegan)

# Ordination on relative abundance (Bray-Curtis is standard)
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))

# Bray-Curtis PCoA (recommended first choice)
ord_pcoa <- ordinate(ps_rel, method = "PCoA", distance = "bray")

# NMDS (for comparison)
ord_nmds <- ordinate(ps_rel, method = "NMDS", distance = "bray", k=2, trymax=100)

# === PCoA Plot ===
p_pcoa <- plot_ordination(ps_rel, ord_pcoa, color = "sample", shape = "sample") +   # adjust variables
  geom_point(size = 4) +
  theme_bw() +
  labs(title = "PCoA - Bray-Curtis Dissimilarity") +
  stat_ellipse(aes(group = sample), level = 0.95)   # optional ellipses

print(p_pcoa)

# === NMDS Plot ===
p_nmds <- plot_ordination(ps_rel, ord_nmds, color = "sample", shape = "sample") +
  geom_point(size = 4) +
  theme_bw() +
  labs(title = "NMDS - Bray-Curtis") +
  stat_ellipse(aes(group = sample), level = 0.95)

print(p_nmds)

# Save plots
ggsave("figures/beta-pcoa-bray.png", p_pcoa, width = 8, height = 6, dpi = 300)
ggsave("figures/beta-nmds-bray.png", p_nmds, width = 8, height = 6, dpi = 300)

# === Distance Matrix (fixed) ===
dist_bray <- phyloseq::distance(ps_rel, method = "bray")

# === PERMANOVA ===
cat("\n=== PERMANOVA (Bray-Curtis) ===\n")
adonis_result <- adonis2(dist_bray ~ sample, 
                         data = as(sample_data(ps), "data.frame"), 
                         permutations = 999)
print(adonis_result)
```

# Functional Groups Analysis - GENERA

```{r ghg-guilds-stacked-100pct, fig.width=11, fig.height=7}
# === Create genus-level relative abundance object ===
ps_genus <- readRDS("output/ps-genus.rds")
ps_genus_rel <- ps_genus %>% 
  transform_sample_counts(function(x) 100 * x / sum(x))

functional_groups <- list(
 
  # === CH4 PRODUCTION (Methanogens) ===
  "CH4_Production" = c(
    "Methanobacterium", "Methanosaeta", "Methanothrix", "Methanosarcina",
    "Methanoculleus", "Methanomassiliicoccus", "Methanoregula",
    "Methanospirillum", "Methanocorpusculum", "Methanofollis", "Methanoplanus",
    "Candidatus Methanofastidiosum"
  ),
 
  # === CH4 CONSUMPTION (Methanotrophs) ===
  "CH4_Consumption" = c(
    "Methylococcus", "Methylosinus", "Methylocystis", "Methylomonas",
    "Methylobacter", "Methylocella", "Methylocapsa",
    "Methylacidiphilum",
    "Methylomirabilis",
    "Candidatus Methanoperedens"
  ),
 
  # === CO2 AEROBIC ===
  "CO2_Aerobic" = c(
    "Mycobacterium", "Pseudomonas", "Sphingomonas", "Burkholderia", "Streptomyces",
    "Flavobacterium", "Variovorax", "Rhizobium", "Paraburkholderia",
    "Candidatus Koribacter", "Candidatus Solibacter"
  ),
 
  # === CO2 ANAEROBIC (Fermenters + Syntrophs + Anaerobes) ===
  "CO2_Anaerobic" = c(
    "Clostridium", "Bacteroides",
    "Syntrophobacter", "Syntrophus", "Pelotomaculum", "Smithella",
    "Geobacter",
    "Desulfovibrio", "Desulfobacter", "Desulfotomaculum",
    "Desulfuromonas", "Shewanella", "Anaeromyxobacter"
  ),
 
  # === NITRIFICATION ===
  "Nitrification" = c(
    "Nitrospira", "Nitrosomonas", "Nitrobacter", "Nitrosospira",
    "Candidatus Nitrosopolaris"
  ),
 
  # === DENITRIFICATION ===
  "Denitrification" = c(
    "Denitratisoma", "Pseudomonas", "Paracoccus", "Thauera",
    "Azoarcus", "Zoogloea", "Dechloromonas"
  ),
 
  # === N-FIXATION ===
  "N_Fixation" = c(
    "Bradyrhizobium", "Rhizobium", "Burkholderia", "Paraburkholderia",
    "Azospirillum", "Azotobacter", "Frankia", "Herbaspirillum",
    "Candidatus Celerinatantimonas", "Candidatus Koribacter"
  )
)

# Summarize + add "Other" category so bars = 100%
guild_summary <- ps_genus_rel %>%
  psmelt() %>%
  mutate(Guild = case_when(
    Genus %in% functional_groups$CH4_Production      ~ "CH4 Production",
    Genus %in% functional_groups$CH4_Consumption     ~ "CH4 Consumption",
    Genus %in% functional_groups$CO2_Aerobic         ~ "CO2 Aerobic",
    Genus %in% functional_groups$CO2_Anaerobic       ~ "CO2 Anaerobic",
    Genus %in% functional_groups$Nitrification ~ "Nitrification",
    Genus %in% functional_groups$Denitrification  ~ "Denitrification",
    Genus %in% functional_groups$N_Fixation    ~ "N-Fixation",
    TRUE ~ "Other"
  )) %>%
  filter(Guild != "Other") %>% 
    mutate(Site = case_when(
    Sample == "Jack001-WW-1" ~ "Wd-2",   # West Woodlands
    Sample == "Jack007-BP-1" ~ "Pr-1",   # Big Prairie
    Sample == "Jack013-SW-1" ~ "Wt-2",   # South Wetlands
    Sample == "Jack019-EW-1" ~ "Wd-1",   # East Woodlands
    Sample == "Jack022-NW-1" ~ "Wt-1",   # North Woodlands?
    Sample == "Jack025-SP-1" ~ "Pr-2",   # South Prairie
    TRUE ~ "Unknown"                     # safety net
  )) %>% 
  group_by(Site, Guild) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Site) %>%
  mutate(Abundance = 100 * Abundance / sum(Abundance))   # ← Normalize to 100%

# === Reorder Sites by increasing CH4 Consumption ===
ch4_order <- guild_summary %>%
  filter(Guild == "CO2 Anaerobic") %>%
  arrange(Abundance) %>%           # Change to desc(Abundance) if you want decreasing
  pull(Site)

guild_order <- c("CH4 Consumption", "CO2 Aerobic", "Nitrification", "Denitrification", "N-Fixation", "CO2 Anaerobic", "CH4 Production")

guild_summary <- guild_summary %>%
  mutate(Site = factor(Site, levels = ch4_order),
         Guild = factor(Guild, levels = guild_order))

# Stacked 100% barplot
p_stacked_100 <- guild_summary %>%
  ggplot(aes(x = Site, y = Abundance, fill = Guild)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  theme(axis.text.x = element_text(hjust = 1, size = 11),
        legend.position = "right",
        legend.text = element_text(size = 9)) +
  labs(title = "Scaled Functional Guild Composition (Genus Level)",
       subtitle = "Relative abundance of key GHG-related microbial guilds",
       y = "Relative Abundance (%)", 
       x = "") +
  scale_fill_viridis_d(option = "turbo")

print(p_stacked_100)

ggsave("figures/ghg_guilds_genera_stacked_100pct.png", p_stacked_100, width = 9, height = 5, dpi = 300)
```

## Dominant genera

```{r}
# Calculate dominant genera
ps_genus_rel %>%
  psmelt() %>%
  mutate(Guild = case_when(
    Genus %in% functional_groups$CH4_Production      ~ "CH4 Production",
    Genus %in% functional_groups$CH4_Consumption     ~ "CH4 Consumption",
    Genus %in% functional_groups$CO2_Aerobic         ~ "CO2 Aerobic",
    Genus %in% functional_groups$CO2_Anaerobic       ~ "CO2 Anaerobic",
    Genus %in% functional_groups$Nitrification ~ "Nitrification",
    Genus %in% functional_groups$Denitrification  ~ "Denitrification",
    Genus %in% functional_groups$N_Fixation    ~ "N-Fixation",
    TRUE ~ "Other"
  )) %>%
  filter(Guild != "Other") %>% 
  group_by(Sample, Guild, Genus) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Sample, Guild) %>%
  mutate(Percent = Abundance / sum(Abundance) * 100) %>%
  slice_max(Abundance) %>% 
  arrange(Guild)
```

## Percent Other

```{r}
# Calculate % Other (and all guilds) properly
guild_percent <- ps_genus_rel %>%
  psmelt() %>%
  mutate(Guild = case_when(
    Genus %in% functional_groups$CH4_Production      ~ "CH4 Production",
    Genus %in% functional_groups$CH4_Consumption     ~ "CH4 Consumption",
    Genus %in% functional_groups$CO2_Aerobic         ~ "CO2 Aerobic",
    Genus %in% functional_groups$CO2_Anaerobic       ~ "CO2 Anaerobic",
    Genus %in% functional_groups$Nitrification ~ "Nitrification",
    Genus %in% functional_groups$Denitrification  ~ "Denitrification",
    Genus %in% functional_groups$N_Fixation    ~ "N-Fixation",
    TRUE ~ "Other"
  )) %>%
  group_by(Sample, Guild) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Sample) %>%
  mutate(Percent = Abundance / sum(Abundance) * 100) %>%
  ungroup()

# View % Other per sample
guild_percent %>%
  filter(Guild == "Other") %>%
  select(Sample, Guild, Percent)

# Optional: Get overall % Other across all samples
guild_percent %>%
  filter(Guild == "Other") %>%
  summarise(Mean_Percent_Other = mean(Percent))
```


# Functional Groups Analysis - FAMILY

```{r ghg-guilds-stacked-100pct-family, fig.width=11, fig.height=7}
# === Create genus-level relative abundance object ===
ps_family <- readRDS("output/ps-family.rds")
ps_family_rel <- ps_family %>% 
  transform_sample_counts(function(x) 100 * x / sum(x))

functional_groups <- list(
 
  # === CH4 PRODUCTION (Methanogens) ===
  "CH4_Production" = c(
    "Methanobacteriaceae",
    "Methanosaetaceae", "Methanotrichaceae",
    "Methanosarcinaceae",
    "Methanomicrobiaceae",
    "Methanomassiliicoccaceae",
    "Methanoregulaceae"
  ),
 
  # === CH4 CONSUMPTION (Methanotrophs) ===
  "CH4_Consumption" = c(
    "Methylococcaceae",
    "Methylocystaceae",
    "Methylacidiphilaceae",
    "Methylomirabilaceae"
  ),
 
  # === CO2 AEROBIC ===
  "CO2_Aerobic" = c(
    "Mycobacteriaceae",
    "Pseudomonadaceae",
    "Sphingomonadaceae",
    "Burkholderiaceae",
    "Streptomycetaceae",
    "Flavobacteriaceae",
    "Gemmatimonadaceae",
    "Acidobacteriaceae"
  ),
 
  # === CO2 ANAEROBIC (Fermenters + Syntrophs + Anaerobes) ===
  "CO2_Anaerobic" = c(
    "Clostridiaceae",
    "Bacteroidaceae",
    "Syntrophobacteraceae",
    "Syntrophaceae",
    "Pelotomaculaceae",
    "Smithellaceae",
    "Geobacteraceae",
    "Desulfovibrionaceae",
    "Desulfuromonadaceae",
    "Desulfobacteraceae",
    "Desulfotomaculaceae",
    "Shewanellaceae",
    "Myxococcaceae"
  ),
 
  # === NITRIFICATION ===
  "Nitrification" = c(
    "Nitrospiraceae",
    "Nitrosomonadaceae",
    "Nitrobacteraceae"
  ),
 
  # === DENITRIFICATION ===
  "Denitrification" = c(
    "Rhodocyclaceae",
    "Pseudomonadaceae",
    "Burkholderiaceae"
  ),
 
  # === N-FIXATION ===
  "N_Fixation" = c(
    "Bradyrhizobiaceae",
    "Rhizobiaceae",
    "Beijerinckiaceae",
    "Xanthobacteraceae",
    "Burkholderiaceae"
  )
)
# === Summarize with "Other" category ===
guild_summary <- ps_family_rel %>%
  psmelt() %>%
mutate(Guild = case_when(
    Family %in% functional_groups$CH4_Production     ~ "CH4 Production",
    Family %in% functional_groups$CH4_Consumption    ~ "CH4 Consumption",
    Family %in% functional_groups$CO2_Aerobic        ~ "CO2 Aerobic",
    Family %in% functional_groups$CO2_Anaerobic ~ "CO2 Anaerobic",
    Family %in% functional_groups$Nitrification      ~ "Nitrification",
    Family %in% functional_groups$Denitrification    ~ "Denitrification",
    Family %in% functional_groups$N_Fixation         ~ "N-Fixation",
    TRUE ~ "Other"
  )) %>%
    filter(Guild != "Other") %>% 
    mutate(Site = case_when(
    Sample == "Jack001-WW-1" ~ "Wd-2",   # West Woodlands
    Sample == "Jack007-BP-1" ~ "Pr-1",   # Big Prairie
    Sample == "Jack013-SW-1" ~ "Wt-2",   # South Wetlands
    Sample == "Jack019-EW-1" ~ "Wd-1",   # East Woodlands
    Sample == "Jack022-NW-1" ~ "Wt-1",   # North Woodlands?
    Sample == "Jack025-SP-1" ~ "Pr-2",   # South Prairie
    TRUE ~ "Unknown"                     # safety net
  )) %>% 
  group_by(Site, Guild) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Site) %>%
  mutate(Abundance = 100 * Abundance / sum(Abundance))   # Normalize to 100%

# === Reorder Sites by increasing CO2 Aerobic ===
ch4_order <- guild_summary %>%
  filter(Guild == "CH4 Consumption") %>%
  arrange(Abundance) %>%           # Change to desc(Abundance) if you want decreasing
  pull(Site)

guild_order <- c("CH4 Consumption", "CO2 Aerobic", "Nitrification", "Denitrification", "N-Fixation", "CO2 Anaerobic", "CH4 Production")

guild_summary <- guild_summary %>%
  mutate(Site = factor(Site, levels = ch4_order),
         Guild = factor(Guild, levels = guild_order))

# === Stacked 100% barplot with reordered x-axis ===
p_stacked_100 <- guild_summary %>%
  ggplot(aes(x = Site, y = Abundance, fill = Guild)) +
  geom_bar(stat = "identity", position = "stack") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
        legend.position = "right",
        legend.text = element_text(size = 9)) +
  labs(title = "Functional Guild Composition (Family Level)",
       subtitle = "Sites ordered by increasing CH4 Consumption",
       y = "Relative Abundance (%)",
       x = "") +
  scale_fill_viridis_d(option = "turbo")

print(p_stacked_100)

ggsave("figures/ghg_guilds_family_stacked_100pct.png", 
       p_stacked_100, width = 9, height = 5, dpi = 300)
```

## Dominant families

```{r}
# Calculate dominant family
ps_family_rel %>%
  psmelt() %>%
mutate(Guild = case_when(
    Family %in% functional_groups$CH4_Production     ~ "CH4 Production",
    Family %in% functional_groups$CH4_Consumption    ~ "CH4 Consumption",
    Family %in% functional_groups$CO2_Aerobic        ~ "CO2 Aerobic",
    Family %in% functional_groups$CO2_Anaerobic ~ "CO2 Anaerobic",
    Family %in% functional_groups$Nitrification      ~ "Nitrification",
    Family %in% functional_groups$Denitrification    ~ "Denitrification",
    Family %in% functional_groups$N_Fixation         ~ "N-Fixation",
    TRUE ~ "Other"
  )) %>%
  filter(Guild != "Other") %>% 
  group_by(Sample, Guild, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Sample, Guild) %>%
  mutate(Percent = Abundance / sum(Abundance) * 100) %>%
  slice_max(Abundance) %>% 
  arrange(Guild)
```

## Parse CO2 Aerobic Families

```{r}
# Calculate dominant family within CO2 Aerobic
ps_family_rel %>%
  psmelt() %>%
  mutate(Guild = case_when(
    Family %in% functional_groups$CH4_Production     ~ "CH4 Production",
    Family %in% functional_groups$CH4_Consumption    ~ "CH4 Consumption",
    Family %in% functional_groups$CO2_Aerobic        ~ "CO2 Aerobic",
    Family %in% functional_groups$CO2_Anaerobic ~ "CO2 Anaerobic",
    Family %in% functional_groups$Nitrification      ~ "Nitrification",
    Family %in% functional_groups$Denitrification    ~ "Denitrification",
    Family %in% functional_groups$N_Fixation         ~ "N-Fixation",
    TRUE ~ "Other"
  )) %>%
  filter(Guild == "CO2 Aerobic") %>% 
  group_by(Sample, Guild, Family) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Sample, Guild) %>%
  mutate(Percent = Abundance / sum(Abundance) * 100) %>%
  slice_max(Abundance, n = 3) %>% 
  arrange()
```

## Percent Other

```{r}
# === Calculate % Other (Family level) ===
guild_percent <- ps_family_rel %>%
  psmelt() %>%
  mutate(Guild = case_when(
    Family %in% functional_groups$CH4_Production     ~ "CH4 Production",
    Family %in% functional_groups$CH4_Consumption    ~ "CH4 Consumption",
    Family %in% functional_groups$CO2_Aerobic        ~ "CO2 Aerobic",
    Family %in% functional_groups$CO2_Anaerobic ~ "CO2 Anaerobic",
    Family %in% functional_groups$Nitrification      ~ "Nitrification",
    Family %in% functional_groups$Denitrification    ~ "Denitrification",
    Family %in% functional_groups$N_Fixation         ~ "N-Fixation",
    TRUE ~ "Other"
  )) %>%
  group_by(Sample, Guild) %>%
  summarise(Abundance = sum(Abundance), .groups = "drop") %>%
  group_by(Sample) %>%
  mutate(Percent = Abundance / sum(Abundance) * 100) %>%
  ungroup()

# View % Other per sample
guild_percent %>%
  filter(Guild == "Other") %>%
  select(Sample, Guild, Percent)

# Get overall average % Other across all samples
guild_percent %>%
  filter(Guild == "Other") %>%
  summarise(Mean_Percent_Other = mean(Percent))
```

